Stochastic blockmodeling of relational event dynamics

### 2012-06-01 docs/abstract.tex
Small edits.

\begin{abstract}
Observing network data as a stream of relational events provides an opportunity to understand differences in how individuals interact over time.
For continuous-time network data, several approaches have recently been proposed for modeling dyadic event rates, conditioned on the observed history of events and nodal or dyadic covariates.  In many cases, however, interaction propensities -- and even the underlying mechanisms of interaction -- vary systematically across subgroups whose identities are unobserved.
For static networks, such heterogeneity has been treated via methods such as stochastic blockmodeling, which operate by assuming latent groups of individuals with similar tendencies in their group-wise interactions.
Here, we combine these two approaches by positing a latent partition on the vertex set, such that event dynamics within and between node sets evolve in potentially distinct ways.
We illustrate the use of our model family by application to several forms of dyadic interaction data, including email communication and Twitter direct messages.
Parameter estimates from the fitted models clearly reveal heterogeneity in the dynamics among groups of individuals.
We also find that the fitted models have better predictive accuracy than either baseline models or relational event models without latent structure.
Our approach illustrates the utility of latent structure methods based on detailed dynamics, which can succeed even in the absence of differences in marginal interaction rates across groups. 
%Our approach illustrates the efficacy of combining a detailed model for local dependencies and a latent variable model for meso-scale dependencies.
\end{abstract}

>>>>

\begin{abstract}
Observing network data as a stream of relational events provides an opportunity to understand differences in how individuals interact over time.
For continuous-time network data, several approaches have recently been proposed for modeling dyadic event rates, conditioned on the observed history of events and nodal or dyadic covariates.  In many cases, however, interaction propensities -- and even the underlying mechanisms of interaction -- vary systematically across subgroups whose identities are unobserved.
For static networks, such heterogeneity has been treated via methods such as stochastic blockmodeling, which operate by assuming latent groups of individuals with similar tendencies in their group-wise interactions.
Here, we combine these two approaches by positing a latent partition of the vertex set such that event dynamics within and between node sets evolve in potentially distinct ways.
We illustrate the use of our model family by application to several forms of dyadic interaction data, including email communication and Twitter direct messages.
Parameter estimates from the fitted models clearly reveal heterogeneity in the dynamics among groups of individuals.
We also find that the fitted models have better predictive accuracy than either baseline models or relational event models without latent structure.
Our approach illustrates the utility of latent structure methods based on detailed dynamics, which can succeed even in the absence of differences in marginal interaction rates across groups. 
% TODO: This last sentence is a bit unclear for a new reader.
\end{abstract}

### 2012-06-01 docs/intro.tex
Small edits.

Borrowing from the intuition of stochastic blockmodels, we propose a continuous-time model of network-based event sequences where latent clusters of nodes share similar patterns of interaction.
Our approach employs a flexible framework for  specifying how the process depends on the previous history of events \cite{AalenOddO.2008, Butts2008}.
%extend continuous-time models from event history analysis to model sequences of network-based event data \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}. % \cite{AalenOddO.2008} 
%These models allow one to specify how the process depends on the previous history of events. % TODO use 'flexible'?
In this way one can investigate theories about the underlying processes and make predictions about future data conditioned on the past.

We describe parameter estimation and learning of  latent cluster assignments via MCMC, and illustrate the behavior of the model with simulated data.
Using several real-world social network data sets involving dyadic communication, we compare the predictive performance of the fitted models to standard baselines.
Finally we show that the parameter estimates exhibit an interpretable structure to the event dynamics, enabling the study of a dynamic extension to stochastic equivalence.
%allowing identification of particular \emph{roles} in  that enable

>>>>

Borrowing from the intuition of stochastic blockmodels, we propose a novel continuous-time model of network-based event sequences where latent clusters of nodes share similar patterns of interaction.
Our approach employs a flexible framework for  specifying how the process depends on the previous history of events \cite{AalenOddO.2008, Butts2008}.
In this way one can compare theories about the underlying processes and make predictions about future data conditioned on the past.

We describe how we learn the latent cluster assignments and model parameters via MCMC, and illustrate the behavior of the model with simulated data.
Using several real-world social network data sets involving dyadic communication, we compare the predictive performance of the fitted models to standard baselines.
We conclude by showing that the parameter estimates reveal interpretable structure in the event dynamics, enabling the study of a dynamic extension to stochastic equivalence.

### 2012-06-01 docs/model.tex
Small edits.

Consider a sequence of events $\mathcal{A} = (0,t_1, \ldots, t_M)$ arising from a nonhomogeneous Poisson process with  intensity $\lambda(t)$.
If the intensity is left continuous and piecewise constant with respect to a set of knots $\tau$ then the likelihood can be written
\begin{align}
\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda(t_m) \exp\left\{ - \int_{0}^{t_M} \lambda(s)ds \right\} \notag  \\
&= \prod_{m=1}^M \lambda(t_m) \prod_{k=1}^{|\tau|} \exp\left\{ - (\tau_{k} - \tau_{k-1}) \lambda(\tau_k) \right\}%  \exp\left\{ - (t - \tau_{M}) \lambda(t) \right\}
\end{align}

One can extend the above to marked point processes where each event contains additional information.
In the context of relational events occurring among $N$ nodes, each event in the process contains both a sender $i$ and a recipient $j$ such that  $(i,j) \in \mathcal{R}$, where $\mathcal{R}$ is the \emph{risk set} comprised of all allowed dyads.
We assume each dyad $(i,j) \in \mathcal{R}$ occurs as a Poisson process with intensity $\lambda_{ij}(t|\cdot)$ that is piecewise constant and depends on the previous history of events  $\mathcal{A}_t = \{(t_m,i_m,j_m): t_m \in [0,t) \}$.
The likelihood of an observed event history $\mathcal{A}_{t_M}$ (also denoted $\mathcal{A}$ for convenience) is
\begin{align}
\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}}\exp\{ - (t_m - t_{m-1}) \lambda_{ij}(t_m | \cdot) \}
\label{eqn:llk}
\end{align}

@@

% labels
\node at (-.5,2) {$\lambda_{1,2}$};
\node at (-.5,1) {$\lambda_{3,4}$};
\node at (1,-.3) {$t_1$};
\node at (3,-.3) {$t_2$};
\end{tikzpicture}
\caption{Intensities for two dyads}
\end{minipage}
\caption[]{Illustration of event data and the assumptions of the model.  Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.  Right: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).  We assume $\lambda_{34}$ can only change after events sent or received by node 3.\footnotemark }
\label{fig:example}
\end{figure}

We aim to learn about the dynamics between subsets of nodes.   To facilitate this, we assume each node $i$ is assigned a latent cluster $z_i$ and use a log linear model for the intensity functions
\begin{align*}
\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{\beta},\mathbf{z}) &= \boldsymbol{\beta}'_{z_i,z_j} \mathbf{s}(t,i,j,\mathcal{A}_t)
%z_i &\sim \mbox{CRP}(\alpha) &
%\boldsymbol{\beta}_{k,l} &\sim %\mbox{N}_p(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)
\end{align*}
where for each pair of clusters $(k,l)$ we have a vector of parameters $\boldsymbol{\beta}_{k,l} \sim \mbox{N}_P(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)$ that corresponds to the vector of $P$ statistics $\mathbf{s}(t,i,j,\mathcal{A}_t)$ computed from the previous history $\mathcal{A}_t$.
Thus, our model for the rate of $(i,j)$ events has the same parameters as for other dyads occurring between group $z_i$ and $z_j$.

 In practice we may believe an interaction between one dyad should not affect the rate of an entirely separate dyad.
As shown in Figure \ref{fig:example}, we allow each intensity function $\lambda_{ij}(t)$ to only change following an event where $i$ was the sender or the recipient.
This is sensible in distributed settings where $i$ has limited knowledge about interactions among other actors.
In the Supplementary Material we show how this assumption also reduces the computational complexity of computing the above likelihood.
%\footnote{TODO: This is analogous to a Markov condition for this class of models.}

We allow the blocks to share information by placing a hierarchical prior on the collection of $\boldsymbol{\beta}_{k.l}$ where $\mu_p \propto 1$ and $\sigma_p^2 \sim \mbox{Inv-Gamma}(\alpha_{\sigma},\beta{\sigma})$.
The cluster assignments are given a non-parameteric priors $z_i \sim \mbox{CRP}(\alpha)$.

@@

\begin{table}[t]
\footnotesize
\center
\begin{tabular}{|l|l|}
\hline
Statistic & Formula \\
\hline
\hline
Intercept& $s_{0}(t,i,j) = 1$\\
Reciprocity (AB-BA)& $s_{1}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$\\
Turn-continuing (AB-AY)& $s_{2}(t,i,j) =  I(i_m=i,j_m=j,i_{v_{mij}}=i,j_{v_{mij}}\ne j)$\\
Turn-taking (AB-BY)&$s_{3}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}\ne i)$\\
%Turn-usurping (AB-XA)& $s_{7}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne j,j_{v_{mij}}=i)$\\
%Turn-usurping (AB-XB)&$s_{8}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne i,j_{v_{mij}}=j)$\\
Sender out-degree& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$\\
Sender in-degree& $s_{5}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=i) )$\\
Dyad count& $s_{6}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i,j_m=j) )$\\
%Receiver out-degree& $s_{3}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=j))$\\
%Receiver in-degree& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=j))$\\
\hline
\end{tabular}
\label{tab:stats}
\caption{Statistics used to specify intensity functions using the previous history $\mathcal{A}_t$.}
\end{table}

In Table \ref{tab:stats} we list the statistics we use in our experiments for  $\mathbf{s}(t,i,j)$.
For example, a particular node sending often may indicate they will continue to be a sender in the future.
We normalize these counts by the number of events up until a dyad's prior changepoint.%\footnote{This seems to avoid ``explosion'' when simulating from the model.  TODO: Explore the theory about these sorts of statistics.  Aalen has a few suggestions.}
Other statistics could be of interest for particular substantive questions \cite{Butts2008,Vu2011}.  \footnote{Describe the form of $f()$.  Right now it is $f(x) = \log \frac{x+1}{m + N(N-1)}$.}

The next set of statistics are \emph{participation shift} effects inspired by research in conversational norms \cite{Gibson2003}.
For example, an \texttt{ab-ba} effect indicates an increased propensity for reciprocity where the event $(a,b)$ is followed by $(b,a)$.
Though we use only the above statistics in our experiments, one may use any quantity that is computed using the previous history of events or known covariates about nodes or dyads.
The only restriction is that the statistic may not change in value between each observed event.\footnote{This prevents one from using the number of events occurring in some previous time window, as in \cite{Gunawardana2011}.}

@@

Our formulation is reminiscent of the stochastic blockmodel \cite{Nowicki2001,Kemp} which models the probability of a dyad as $p(y_{ij}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is interpreted as a mixing rate between group $z_i$ and group $z_j$.
In the proposed method, however, the blockmodel structure facilitates the study of intra-group and inter-group dynamics via a continuous-time network model.

The above framework generalizes several important special cases.
For example,  using only the intercept statistic $s_0(t,i,j) = 1$ is analogous to the stochastic block model for static networks.
Under this model each dyad is a homogeneous Poisson process and all dyad intensities $\lambda_{i,j}$ within block $(z_i,z_j)$ have the same intensity, $\exp\{\boldsymbol{\beta}_{z_i,z_j}\}$.
As these intensities do not change under this specification, the likelihood simplifies to
$$\mathcal{L}(\mathcal{A}|\beta) = \prod_{m=1}^M \lambda_{i_m,j_m} \prod_{(i,j) \in \mathcal{R}} \exp\{-t_M \lambda_{i,j}\}$$

@@

\subsection{Related work}
The proposed model is an extension of recent work on modeling event-based social network data using an event history approach \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}. % \cite{AalenOddO.2008} 
Relational event models such as these require a knot at each observed event, while other approaches such as \cite{Gunawardana2011} learn the regions where an intensity is constant.
By using decision trees,  \cite{Gunawardana2011} also allow for a nonlinear relationship between statistics and intensity functions, while here we use latent variables to allow for heterogeneity in intensities across possible events.

>>>>

Consider a sequence of events $\mathcal{A} = (0,t_1, \ldots, t_M)$ arising from a nonhomogeneous Poisson process with  intensity $\lambda(t)$.
If the intensity is left continuous and piecewise constant with respect to a set of knots $\tau$ then the likelihood can be written
\begin{align}
\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda(t_m) \exp\left\{ - \int_{0}^{t_M} \lambda(s)ds \right\} \notag  \\
&= \prod_{m=1}^M \lambda(t_m) \prod_{k=1}^{|\tau|} \exp\left\{ - (\tau_{k} - \tau_{k-1}) \lambda(\tau_k) \right\}%  \exp\left\{ - (t - \tau_{M}) \lambda(t) \right\}
\end{align}
%{\color{red} (You've left out the right-censoring factor here (and also below).)}

One can extend the above to marked point processes where each event contains additional information.
In the context of relational events occurring among $N$ nodes, each event in the process contains both a sender $i$ and a recipient $j$ such that  $(i,j) \in \mathcal{R}$, where $\mathcal{R}$ is the \emph{risk set} comprised of all allowed dyads.
We assume each event involving dyad $(i,j) \in \mathcal{R}$ occurs as a Poisson process with intensity $\lambda_{ij}(t|\cdot)$ that is piecewise constant and depends on the previous history of events  $\mathcal{A}_t = \{(t_m,i_m,j_m): t_m \in [0,t) \}$.
The likelihood of an observed event history $\mathcal{A}_{t_M}$ (also denoted $\mathcal{A}$ for convenience) is
\begin{align}
\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}}\exp\{ - (t_m - t_{m-1}) \lambda_{ij}(t_m | \cdot) \}
\label{eqn:llk}
\end{align}

@@

% labels
\node at (-.5,2) {$\lambda_{1,2}$};
\node at (-.5,1) {$\lambda_{3,4}$};
\node at (1,-.3) {$t_1$};
\node at (3,-.3) {$t_2$};
\end{tikzpicture}
\caption{Intensities for two dyads}
\end{minipage}
\caption[]{Illustration of event data and the assumptions of the model.  Left: A sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.  Right: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).  We assume $\lambda_{34}$ can only change after events sent or received by node 3.\footnotemark }
\label{fig:example}
\end{figure}

We aim to learn about the dynamics between subsets of nodes.   To facilitate this, we assume each node $i$ belongs to latent cluster $z_i$ and use a log linear model for the intensity functions
\begin{align*}
\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{\beta},\mathbf{z}) &= \boldsymbol{\beta}'_{z_i,z_j} \mathbf{s}(t,i,j,\mathcal{A}_t)
%z_i &\sim \mbox{CRP}(\alpha) &
%\boldsymbol{\beta}_{k,l} &\sim %\mbox{N}_p(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)
\end{align*}
where for each pair of clusters $(k,l)$ we have a parameter vector $\boldsymbol{\beta}_{k,l} \sim \mbox{N}_P(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)$ that corresponds to the $P$-dimensional vector of statistics $\mathbf{s}(t,i,j,\mathcal{A}_t)$ computed from the previous history $\mathcal{A}_t$.
Thus, our model for the rate of $(i,j)$ events has the same parameters as for other events occurring between group $z_i$ and $z_j$.

 In practice we may believe an interaction between one dyad should not affect the rate of an entirely separate dyad.
As shown in Figure \ref{fig:example}, we only allow each intensity function $\lambda_{ij}(t)$ to change following an event where $i$ was the sender or the recipient.
This is sensible in distributed settings where $i$ has limited knowledge about interactions among other actors.
In the Supplementary Material we show how this assumption also reduces the computational complexity of computing the above likelihood.
% TODO: Lead off with computational convenience, then justify?
%\footnote{TODO: This is analogous to a Markov condition for this class of models.}

We allow the blocks to share information by placing a hierarchical prior on the collection of $\boldsymbol{\beta}_{k.l}$ where $\mu_p \propto 1$ and $\sigma_p^2 \sim \mbox{Inv-Gamma}(\alpha_{\sigma},\beta{\sigma})$.
The cluster assignments are given a non-parameteric prior $z_i \sim \mbox{CRP}(\alpha)$.
% TODO: Expand on this?

@@

\begin{table}[t]
\footnotesize
\center
\begin{tabular}{|l|l|}
\hline
Statistic & Formula\\
\hline
\hline
Intercept& $s_{0}(t,i,j) = 1$\\
Reciprocity (AB-BA)& $s_{1}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$\\
Turn-continuing (AB-AY)& $s_{2}(t,i,j) =  I(i_m=i,j_m=j,i_{v_{mij}}=i,j_{v_{mij}}\ne j)$\\
Turn-taking (AB-BY)&$s_{3}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}\ne i)$\\
%Turn-usurping (AB-XA)& $s_{7}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne j,j_{v_{mij}}=i)$\\
%Turn-usurping (AB-XB)&$s_{8}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne i,j_{v_{mij}}=j)$\\
Sender out-degree& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$\\
Sender in-degree& $s_{5}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=i) )$\\
Dyad count& $s_{6}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i,j_m=j) )$\\
%Receiver out-degree& $s_{3}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=j))$\\
%Receiver in-degree& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=j))$\\
\hline
\end{tabular}
\label{tab:stats}
\caption{Statistics used to specify intensity functions using the previous history $\mathcal{A}_t$, where $v_{mij}$ is the index of the changepoint for $\lambda_{ij}$ previous to event $m$.}
\end{table}

In Table \ref{tab:stats} we list the statistics we use in our experiments for  $\mathbf{s}(t,i,j)$.
For example, a particular node sending often may indicate they will continue to be a sender in the future.
We normalize these counts by the number of events up until a dyad's prior changepoint. %\footnote{This seems to avoid ``explosion'' when simulating from the model.  TODO: Explore the theory about these sorts of statistics.  Aalen has a few suggestions.}
Other statistics could be of interest for particular substantive questions \cite{Butts2008,Vu2011}.  \footnote{Describe the form of $f()$.  Right now it is $f(x) = \log \frac{x+1}{m + N(N-1)}$.}

% TODO: Awkward next sentence
The next set of statistics are \emph{participation shift} effects inspired by research in conversational norms \cite{Gibson2003}.
For example, an \texttt{ab-ba} effect indicates an increased propensity for reciprocity where the event $(a,b)$ is followed by $(b,a)$.
Though we use only the statistics in Table \ref{tab:stats} in our experiments, one may use any quantity that is computed using the previous history of events or known covariates about nodes or dyads.
The only restriction is that the statistic may not change in value between each observed event.\footnote{This prevents one from using the number of events occurring in some previous time window, as in \cite{Gunawardana2011}.}
% TODO: Also, need to be restricted to ego conditions if we want it fast

@@

Our formulation is reminiscent of the stochastic blockmodel \cite{Nowicki2001,Kemp} for static networkswhich models the probability of a dyad as $p(y_{ij}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is interpreted as a mixing rate between group $z_i$ and group $z_j$.
In the proposed method, however, the blockmodel structure facilitates the study of intra-group and inter-group dynamics via a continuous-time network model.

The proposed family of models generalizes several important special cases.
For example,  using only the intercept statistic $s_0(t,i,j) = 1$ is analogous to the stochastic block model for static networks.
Under this model each dyad is a homogeneous Poisson process and all dyad intensities $\lambda_{i,j}$ within block $(z_i,z_j)$ have the same intensity, $\exp\{\boldsymbol{\beta}_{z_i,z_j}\}$.
As these intensities do not change under this specification, the likelihood simplifies to
$$\mathcal{L}(\mathcal{A}|\beta) = \prod_{m=1}^M \lambda_{i_m,j_m} \prod_{(i,j) \in \mathcal{R}} \exp\{-t_M \lambda_{i,j}\}$$

@@

\subsection{Related work}
The proposed model is an extension of recent work on modeling event-based social network data using an event history approach \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}. % \cite{AalenOddO.2008} 
Relational event models such as these require a knot at each observed event, while other approaches such as \cite{Gunawardana2011} learn the regions where an intensity is constant.
By using decision trees,  \cite{Gunawardana2011} also allow for a nonlinear relationship between statistics and intensity functions.  
We instead use latent variables to allow for heterogeneity in intensities across possible events.
